---
title: "DNN-Derived Synthetic Vegetation Index for Phenotype Prediction"
author: "Siim Sepp"
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

# Overview

This script uses a deep neural network (DNN) to derive **data-driven vegetation
indices** from raw multispectral drone imagery for the NY Winter Malting Barley
breeding programme.

## What it does

Standard vegetation indices (NDVI, NDRE, etc.) are fixed, hand-crafted
combinations of spectral bands designed for generic vegetation monitoring. Here,
we instead **let the data determine the optimal band combination** for each
agronomic or malt-quality trait. The result is a trait-specific **Synthetic VI**
— a scalar value computed from the five raw bands (blue, green, red, red_edge,
nir) that maximises predictive accuracy for that trait.

A separate model is trained for each phenotype trait, yielding a distinct
Synthetic VI per trait. After model training, the learned encoder weights are
extracted and presented as an explicit mathematical formula so the index can be
computed directly from band reflectances without re-running the DNN.

## Modelling approach

The DNN has a **bottleneck architecture** with two components:

1. **Band Encoder** (5 → 16 → 8 → 1): three fully-connected layers with ReLU
   activations compress the five bands into a single scalar — the Synthetic VI.
   The final layer has no activation, so the index spans the full real line.

2. **Prediction Head** (1 → 8 → 1): maps the Synthetic VI through one hidden
   layer to the trait value.

The bottleneck forces all spectral information through a single neuron before
prediction, mimicking the structure of hand-crafted indices (e.g. NDVI uses a
single derived value) while allowing non-linear, data-optimised band combinations.
Only five specific Timepoint × Environment groups are used, one per trial year,
representing the flight timing judged most informative per environment:
`HELF24_TP7`, `KET21_TP4`, `MCG23_TP5`, `MCG25_TP12`, `SNY22_TP6`.

## Cross-validation

Two complementary CV scenarios are implemented. In both, splits are by **GID**
(genotype ID) to prevent data leakage, and scalers are fit on training data only
then applied to test folds.

### Leave-One-Environment-Out (LOEO)

All observations from one environment are withheld; the model trains on the
remaining four. This measures **cross-environment generalisation**.

MCG25 is excluded from all analyses (see MCG25 Diagnostic section) because its
late-season flight (TP12) produces an inverted spectral–yield relationship
incompatible with the other four environments.

| Fold | Training environments | Validation environment |
|------|-----------------------|------------------------|
| 1 | KET21, MCG23, SNY22 | HELF24 |
| 2 | HELF24, MCG23, SNY22 | KET21 |
| 3 | HELF24, KET21, SNY22 | MCG23 |
| 4 | HELF24, KET21, MCG23 | SNY22 |

### Within-Environment 5-Fold CV (by GID)

Within each environment, GIDs are shuffled and assigned to 5 folds. The model
trains on 4/5 of GIDs and predicts the remaining 1/5. This measures
**within-environment predictive ability** for unseen genotypes.

Each scenario is repeated 3 times with different random seeds. Performance is
Pearson *r* and RMSE on the back-transformed trait scale, benchmarked against
linear regression on standard VIs under the same schemes.

```{r libraries}
# ── Core data wrangling & visualisation ────────────────────────────────────
library(tidyverse)
library(ggplot2)
library(patchwork)
library(corrplot)
library(knitr)
library(kableExtra)

# ── Deep learning ──────────────────────────────────────────────────────────
# torch is the primary backend; neuralnet is a lightweight fallback
if (!requireNamespace("torch", quietly = TRUE)) {
  stop("Package 'torch' is required. Install with: install.packages('torch'); torch::install_torch()")
}
library(torch)
if (!torch::torch_is_installed()) torch::install_torch()
set.seed(42)
torch_manual_seed(42)
```

# Data Loading and Exploration

```{r load-data}
# ── Resolve data path ──────────────────────────────────────────────────────
pheno_paths <- c(
  "C:/Users/Siim Sepp/NY_WMBCL/data/WMB_pheno.Rdata",
  "~/Documents/GitHub/NY_WMBCL/data/WMB_pheno.Rdata",
  "data/WMB_pheno.Rdata"
)
pheno_file <- pheno_paths[file.exists(pheno_paths)][1]
if (is.na(pheno_file)) stop("WMB_pheno.Rdata not found in any expected location.")
message("Loading: ", pheno_file)
load(pheno_file)   # loads one or more objects into the environment

# ── Identify the main data frame ───────────────────────────────────────────
# List objects and find the data frame that is most likely the phenotype table
loaded_objs <- ls()
df_candidates <- loaded_objs[sapply(loaded_objs, function(x) is.data.frame(get(x)))]
message("Data frames found: ", paste(df_candidates, collapse = ", "))

# Use the largest data frame if multiple exist
if (length(df_candidates) == 1) {
  pheno_raw <- get(df_candidates)
} else {
  sizes      <- sapply(df_candidates, function(x) nrow(get(x)))
  pheno_raw  <- get(df_candidates[which.max(sizes)])
  message("Using: ", df_candidates[which.max(sizes)])
}

cat("Dimensions:", nrow(pheno_raw), "rows ×", ncol(pheno_raw), "cols\n")
glimpse(pheno_raw)
```

```{r identify-columns}
# ── Band columns ───────────────────────────────────────────────────────────
# Use anchored / precise patterns so that e.g. "red" does NOT match "red_edge"
BAND_NAMES <- c("blue", "green", "red", "red_edge", "nir")
band_patterns <- c(
  blue     = "^blue$",
  green    = "^green$",
  red      = "^red$",                          # exact – must NOT match red_edge
  red_edge = "red[._]?edge|rededge",
  nir      = "^nir$"
)
band_cols <- sapply(names(band_patterns), function(b) {
  m <- grep(band_patterns[b], names(pheno_raw), ignore.case = TRUE, value = TRUE)
  if (length(m) == 0) NA_character_ else m[1]
}, USE.NAMES = TRUE)
band_cols <- band_cols[!is.na(band_cols)]
# band_cols: names = canonical BAND_NAMES, values = actual column names in pheno_raw
cat("Band columns found:\n"); print(band_cols)
if (length(band_cols) < 5)
  warning("Only ", length(band_cols), " of 5 expected bands found: ",
          paste(setdiff(BAND_NAMES, names(band_cols)), collapse = ", "), " missing.")

# ── Standard VI columns ────────────────────────────────────────────────────
vi_candidates <- c("NDVI","NDRE","EVI","GNDVI","SAVI","ENDVI","GDVI","NRI","GIPVI","PSRI")
vi_cols <- intersect(vi_candidates, names(pheno_raw))
cat("Standard VI columns found:", paste(vi_cols, collapse = ", "), "\n")

# ── Timepoint column ───────────────────────────────────────────────────────
tp_col <- intersect(c("Timepoint","timepoint","TP","tp","time_point"),
                    names(pheno_raw))[1]
cat("Timepoint column:", tp_col, "\n")

# ── Candidate trait columns ────────────────────────────────────────────────
# Exclude IDs, bands, VIs and spatial variables
exclude_pat <- paste(c("GID","plot_id","Env","Row","Col","Timepoint","Fam",
                       "JD","DAS","GDD", BAND_NAMES, vi_candidates),
                     collapse="|")
#trait_cols <- names(pheno_raw)[
#  !grepl(exclude_pat, names(pheno_raw), ignore.case = TRUE) &
#  sapply(pheno_raw, is.numeric)
#]

trait_cols<- colnames(pheno_raw)[19]
cat("Candidate trait columns:\n"); print(trait_cols)

# ── Environment column ─────────────────────────────────────────────────────
env_col <- intersect(c("Env","env","environment","trial"), names(pheno_raw))[1]
cat("Environment column:", env_col, "\n")

# ── GID column ─────────────────────────────────────────────────────────────
gid_col <- intersect(c("GID","gid","Genotype","genotype","EntryName","entry_name",
                       "Line","line","Name","name"),
                     names(pheno_raw))[1]
cat("GID column:", gid_col, "\n")
```

```{r data-summary}
# ── Descriptive summary ────────────────────────────────────────────────────
cat("\nEnvironments present:\n")
print(table(pheno_raw[[env_col]]))

cat("\nBand value ranges:\n")
pheno_raw %>%
  select(all_of(unname(band_cols))) %>%
  summarise(across(everything(), list(
    min  = ~min(., na.rm = TRUE),
    mean = ~mean(., na.rm = TRUE),
    max  = ~max(., na.rm = TRUE),
    pct_na = ~mean(is.na(.)) * 100
  ))) %>%
  pivot_longer(everything(), names_to = c("band","stat"), names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  kable(digits = 3, caption = "Band summary statistics") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

# Data Preparation

```{r data-prep}
# ── Build modelling dataset ────────────────────────────────────────────────
found_band_names <- names(band_cols)

tp_cols_select  <- if (!is.na(tp_col))  tp_col  else character(0)
gid_cols_select <- if (!is.na(gid_col)) gid_col else character(0)

model_data <- pheno_raw %>%
  select(all_of(c(gid_cols_select, env_col, tp_cols_select,
                  unname(band_cols), vi_cols, trait_cols))) %>%
  rename(Env = all_of(env_col)) %>%
  rename(all_of(band_cols)) %>%
  filter(if_all(all_of(found_band_names), ~ !is.na(.))) %>%
  mutate(Env = as.factor(Env))

# ── Standardise GID column name ────────────────────────────────────────────
if (!is.na(gid_col) && gid_col %in% names(model_data) && gid_col != "GID") {
  model_data <- model_data %>% rename(GID = all_of(gid_col))
} else if (!"GID" %in% names(model_data)) {
  model_data$GID <- as.character(seq_len(nrow(model_data)))
  warning("No GID column found; using row index as GID.")
}

# ── Standardise Timepoint column name ─────────────────────────────────────
if (!is.na(tp_col) && tp_col != "Timepoint") {
  model_data <- model_data %>% rename(Timepoint = all_of(tp_col))
}
if (!"Timepoint" %in% names(model_data)) {
  model_data$Timepoint <- 1L
  warning("No Timepoint column found; treating all rows as a single timepoint.")
}

# ── Group variable: Env × Timepoint ───────────────────────────────────────
model_data <- model_data %>%
  mutate(Timepoint = as.integer(Timepoint),
         group     = paste(Env, Timepoint, sep = "_TP"))

model_data <- model_data %>%
  filter(group %in% c("HELF24_TP7","KET21_TP4","MCG23_TP5","MCG25_TP12","SNY22_TP6"))

BAND_NAMES <- found_band_names
cat("\nRows with complete band data (all 5 environments):", nrow(model_data), "\n")

# ── Compute NDVI from raw band values (before any standardisation) ────────
if (all(c("red", "nir") %in% names(model_data))) {
  model_data <- model_data %>%
    mutate(NDVI_raw = (nir - red) / (nir + red + 1e-8))
  cat("NDVI_raw computed: min =", round(min(model_data$NDVI_raw, na.rm=TRUE), 3),
      " mean =", round(mean(model_data$NDVI_raw, na.rm=TRUE), 3),
      " max =", round(max(model_data$NDVI_raw, na.rm=TRUE), 3), "\n")
} else {
  warning("Cannot compute NDVI_raw: 'red' or 'nir' not found in band columns.")
}

# ── Save truly raw copy with all 5 environments (used in MCG25 diagnostic) ─
model_data_raw_orig <- model_data

# ── Exclude MCG25: TP12 is a late-season flight with inverted spectral-yield
#    relationship (see MCG25 Diagnostic section). All model training uses
#    the remaining 4 environments only.
model_data <- model_data %>% filter(Env != "MCG25")
cat("MCG25 excluded. Remaining environments:",
    paste(sort(unique(as.character(model_data$Env))), collapse = ", "), "\n")
cat("Remaining rows:", nrow(model_data), "\n")

group_levels_all <- sort(unique(model_data$group))
cat("Timepoint × Environment groups:", length(group_levels_all), "\n")
print(group_levels_all)

# ── Per-environment spectral standardisation ──────────────────────────────
# Each band is z-scored within its environment, removing environment-specific
# spectral baseline offsets (sensor calibration, soil background, canopy
# structure differences between trials). Applied after NDVI_raw so NDVI
# uses original reflectances. Both model_data_raw (CV) and model_data
# (full-data models) receive these standardised bands.
model_data <- model_data %>%
  group_by(Env) %>%
  mutate(across(all_of(BAND_NAMES),
                ~ (. - mean(., na.rm = TRUE)) / (sd(., na.rm = TRUE) + 1e-8))) %>%
  ungroup()

cat("\nPer-environment band standardisation applied.\n")
cat("Post-standardisation band means per environment (should be ~0):\n")
model_data %>%
  group_by(Env) %>%
  summarise(across(all_of(BAND_NAMES), ~ round(mean(., na.rm = TRUE), 3)),
            .groups = "drop") %>%
  print()

# ── Save env-standardised, unscaled copy for per-fold scaling in CV ────────
model_data_raw <- model_data

# ── Global scaling of model_data (used for full-data models & plots) ───────
band_scale <- map(BAND_NAMES, function(b) {
  list(min = min(model_data[[b]], na.rm = TRUE),
       max = max(model_data[[b]], na.rm = TRUE))
}) %>% set_names(BAND_NAMES)

for (b in BAND_NAMES) {
  rng <- band_scale[[b]]
  model_data[[b]] <- (model_data[[b]] - rng$min) / (rng$max - rng$min + 1e-8)
}

trait_scale <- map(trait_cols, function(t) {
  vals <- model_data[[t]]
  list(mean = mean(vals, na.rm = TRUE), sd = sd(vals, na.rm = TRUE))
}) %>% set_names(trait_cols)

for (t in trait_cols) {
  sc <- trait_scale[[t]]
  model_data[[t]] <- (model_data[[t]] - sc$mean) / (sc$sd + 1e-8)
}

cat("Traits to model:", length(trait_cols), "\n")
```

```{r band-correlations, fig.cap="Correlations among raw spectral bands"}
# ── Visual EDA: band-band and band-VI correlations ─────────────────────────
cor_bands <- cor(model_data[, BAND_NAMES], use = "pairwise.complete.obs")
corrplot(cor_bands, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.8,
         title = "Band–Band Correlations", mar = c(0,0,2,0))
```

# MCG25 Diagnostic Analysis

MCG25 (timepoint 12) is the single environment that consistently fails in LOEO
cross-validation. This section investigates whether the failure is driven by
spectral range differences, a different yield–NDVI relationship within the
environment, or the unusually late flight timing relative to the other four trials.

```{r mcg25-timepoint-context}
# ── Timepoint context: how late is MCG25_TP12 relative to the other groups? ─
tp_context <- model_data_raw_orig %>%
  distinct(Env, Timepoint, group) %>%
  arrange(Timepoint)

tp_context %>%
  kable(col.names = c("Environment", "Timepoint", "Group"),
        caption = "Selected Timepoint × Environment groups — MCG25 uses a much later flight") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r mcg25-yield-ndvi-dist, fig.height=5, fig.cap="Yield and NDVI distributions per environment"}
p_yield <- model_data_raw_orig %>%
  filter(!is.na(.data[[trait_cols[1]]])) %>%
  ggplot(aes(x = .data[[trait_cols[1]]], fill = Env)) +
  geom_density(alpha = 0.65) +
  facet_wrap(~Env, ncol = 5, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = trait_cols[1], y = "Density", title = "Yield distribution") +
  theme_bw(base_size = 10) + theme(legend.position = "none")

p_ndvi <- model_data_raw_orig %>%
  ggplot(aes(x = NDVI_raw, fill = Env)) +
  geom_density(alpha = 0.65) +
  facet_wrap(~Env, ncol = 5, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "NDVI (raw)", y = "Density", title = "NDVI distribution") +
  theme_bw(base_size = 10) + theme(legend.position = "none")

p_yield / p_ndvi
```

```{r mcg25-band-distributions, fig.height=4, fig.cap="Raw band reflectance distributions by environment — MCG25 at TP12 may occupy a different spectral range"}
model_data_raw_orig %>%
  select(Env, all_of(BAND_NAMES)) %>%
  pivot_longer(all_of(BAND_NAMES), names_to = "band", values_to = "value") %>%
  mutate(band = factor(band, levels = BAND_NAMES)) %>%
  ggplot(aes(x = Env, y = value, fill = Env)) +
  geom_boxplot(outlier.size = 0.3, alpha = 0.7) +
  facet_wrap(~band, scales = "free_y", nrow = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Reflectance", title = "Band reflectance by environment") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "none")
```

```{r mcg25-ndvi-yield-scatter, fig.height=6, fig.cap="NDVI vs yield within each environment — correlation and regression line"}
trait_diag <- trait_cols[1]

cors_ndvi <- model_data_raw_orig %>%
  filter(!is.na(.data[[trait_diag]]), !is.na(NDVI_raw)) %>%
  group_by(Env) %>%
  summarise(r     = cor(NDVI_raw, .data[[trait_diag]], use = "complete.obs"),
            label = paste0("r = ", round(r, 3)),
            .groups = "drop")

model_data_raw_orig %>%
  filter(!is.na(.data[[trait_diag]]), !is.na(NDVI_raw)) %>%
  ggplot(aes(x = NDVI_raw, y = .data[[trait_diag]], colour = Env)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.8) +
  geom_label(data = cors_ndvi,
             aes(x = -Inf, y = Inf, label = label),
             hjust = -0.05, vjust = 1.4,
             inherit.aes = FALSE, size = 3.5, fontface = "bold",
             fill = "white", label.size = 0.2) +
  facet_wrap(~Env, ncol = 3, scales = "free") +
  scale_colour_brewer(palette = "Set2") +
  labs(x = "NDVI (raw)", y = trait_diag,
       title = "NDVI vs yield: within-environment scatter") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none")
```

```{r mcg25-band-yield-cors}
# ── Within-environment Pearson r for every band and NDVI vs trait ──────────
trait_diag <- trait_cols[1]

band_yield_cors <- model_data_raw_orig %>%
  filter(!is.na(.data[[trait_diag]])) %>%
  group_by(Env) %>%
  summarise(
    across(all_of(c(BAND_NAMES, "NDVI_raw")),
           ~ round(cor(., .data[[trait_diag]], use = "complete.obs"), 3)),
    n_obs = sum(!is.na(.data[[trait_diag]])),
    .groups = "drop"
  )

band_yield_cors %>%
  kable(caption = paste("Within-environment Pearson r: each band and NDVI vs", trait_diag)) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r mcg25-band-yield-heatmap, fig.height=3.5, fig.cap="Heatmap of within-environment band–yield correlations"}
band_yield_cors %>%
  select(Env, all_of(BAND_NAMES), NDVI_raw) %>%
  pivot_longer(-Env, names_to = "band", values_to = "r") %>%
  mutate(band = factor(band, levels = c(BAND_NAMES, "NDVI_raw"))) %>%
  ggplot(aes(x = band, y = Env, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(r, 2)), size = 3.5, fontface = "bold") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = "Spectral predictor", y = NULL,
       title = paste("Within-environment Pearson r — band/NDVI vs", trait_diag),
       fill = "r") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r mcg25-between-within-variance}
# ── Variance partitioning: η² (between-env SS / total SS) per band ────────
# High η² means most spectral variance is between environments (not genotypes).
# If MCG25 is spectrally extreme, it will dominate the between-env component.
vp <- model_data_raw_orig %>%
  select(Env, all_of(BAND_NAMES), NDVI_raw, all_of(trait_cols[1])) %>%
  pivot_longer(-Env, names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(variable) %>%
  mutate(grand_mean = mean(value)) %>%
  group_by(variable, Env) %>%
  mutate(env_mean = mean(value)) %>%
  ungroup() %>%
  group_by(variable) %>%
  summarise(
    SS_total   = sum((value - grand_mean)^2),
    SS_between = sum((env_mean - grand_mean)^2),
    eta2       = round(SS_between / SS_total, 3),
    .groups = "drop"
  ) %>%
  arrange(desc(eta2))

vp %>%
  kable(col.names = c("Variable", "SS total", "SS between-env", "η² (between-env)"),
        digits = c(0, 0, 0, 3),
        caption = "Variance partitioning: proportion of total variance attributable to environment") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

# DNN Architecture

The network has two components trained **within each Timepoint × Environment group**:

1. **Band Encoder** — three fully-connected layers (n_bands → 16 → 8 → 1).
   The single output neuron is the *Synthetic VI* — no activation, so it spans
   the full real line.

2. **Prediction Head** — maps the scalar Synthetic VI through one hidden layer
   to the trait prediction.

The prediction head receives the Synthetic VI concatenated with a learned
**environment embedding** so that the VI→trait mapping can shift per trial,
while the encoder (VI formula) remains shared across environments.
For unseen environments in LOEO, the mean embedding of the training environments
is used as a proxy.

```{r dnn-definition}
# ══════════════════════════════════════════════════════════════════════════
# Band Encoder: n_bands → 16 → 8 → 1 (Synthetic VI bottleneck)
# ══════════════════════════════════════════════════════════════════════════
BandEncoder <- nn_module(
  classname = "BandEncoder",
  initialize = function(n_bands = 5, h1 = 16, h2 = 8) {
    self$fc1  <- nn_linear(n_bands, h1)
    self$fc2  <- nn_linear(h1, h2)
    self$fc3  <- nn_linear(h2, 1)      # bottleneck = Synthetic VI (linear out)
    self$drop <- nn_dropout(0.1)
  },
  forward = function(x) {
    x <- x %>% self$fc1() %>% nnf_relu() %>% self$drop()
    x <- x %>% self$fc2() %>% nnf_relu() %>% self$drop()
    self$fc3(x)
  }
)

# ══════════════════════════════════════════════════════════════════════════
# Full predictor: encoder → Synthetic VI → prediction head → trait
# ══════════════════════════════════════════════════════════════════════════
SyntheticVIModel <- nn_module(
  classname = "SyntheticVIModel",
  initialize = function(n_bands = 5, n_envs = 5, d_embed = 4L, pred_h = 8L) {
    self$encoder  <- BandEncoder(n_bands)
    self$env_emb  <- nn_embedding(n_envs, d_embed)     # one vector per environment
    self$pred_fc1 <- nn_linear(1L + d_embed, pred_h)   # [SynVI | env_embed] → hidden
    self$pred_fc2 <- nn_linear(pred_h, 1L)
    self$drop     <- nn_dropout(0.1)
  },
  forward = function(bands, env_idx = NULL, env_vec = NULL) {
    vi <- self$encoder(bands)                           # (N, 1)
    ev <- if (!is.null(env_idx)) {
      self$env_emb(env_idx)                             # (N, d_embed) — known env
    } else {
      env_vec$expand(c(vi$size(1), -1L))                # broadcast mean embed for unseen env
    }
    torch_cat(list(vi, ev), dim = 2) %>%
      self$pred_fc1() %>% nnf_relu() %>% self$drop() %>%
      self$pred_fc2()
  },
  get_synthetic_vi = function(bands) {
    self$encoder(bands)
  }
)

cat("Model architecture defined.\n")
cat("  Encoder:       n_bands → 16 → 8 → 1 (Synthetic VI)\n")
cat("  Env embedding: n_envs × d_embed = 4\n")
cat("  Predictor:     (1 + 4) → 8 → 1 (trait)\n")
```

# Training

Two CV scenarios are run in parallel:

- **LOEO**: holds out one environment at a time, trains on the other four.
- **Within-env 5-fold CV**: within each environment, GIDs are assigned to 5 folds;
  the model trains on 4 folds and predicts the 5th.

In both, the band scaler (0–1 min–max) and trait scaler (z-score) are fit on
training data only and applied to the test fold — `model_data_raw` (unscaled) is
passed to the CV functions so scaling is done internally.

After CV, a final model trained on all data is used to extract the **Synthetic VI formula**.

```{r training-functions}
# ── Helper: build band + label tensors from a pre-scaled data frame ────────
make_tensors <- function(df, trait) {
  df_sub <- df[!is.na(df[[trait]]), ]
  list(
    bands = torch_tensor(as.matrix(df_sub[, BAND_NAMES, drop = FALSE]),
                         dtype = torch_float()),
    y     = torch_tensor(df_sub[[trait]], dtype = torch_float())$unsqueeze(2)
  )
}

# ── Scale df using stats from train_df; return scaled df + back-transform params
scale_from_train <- function(df, train_df, trait) {
  band_sc  <- map(BAND_NAMES, function(b)
    list(mn = min(train_df[[b]], na.rm = TRUE),
         mx = max(train_df[[b]], na.rm = TRUE))) %>% set_names(BAND_NAMES)
  trait_mn <- mean(train_df[[trait]], na.rm = TRUE)
  trait_sd <- sd(train_df[[trait]],   na.rm = TRUE)
  for (b in BAND_NAMES) {
    sc <- band_sc[[b]]
    df[[b]] <- (df[[b]] - sc$mn) / (sc$mx - sc$mn + 1e-8)
  }
  df[[trait]] <- (df[[trait]] - trait_mn) / (trait_sd + 1e-8)
  list(df = df, trait_mn = trait_mn, trait_sd = trait_sd)
}

# ── Core training loop ─────────────────────────────────────────────────────
train_model <- function(train_df, trait, env_levels,
                        epochs = 300, lr = 3e-3, batch_size = 64,
                        seed = NULL, verbose = FALSE) {
  if (!is.null(seed)) torch_manual_seed(seed)
  df_sub  <- train_df[!is.na(train_df[[trait]]), ]
  tensors <- make_tensors(train_df, trait)
  n       <- tensors$bands$size(1)
  if (n < 10) return(NULL)

  # Environment indices (1-based, as required by torch nn_embedding in R)
  env_idx <- torch_tensor(
    match(as.character(df_sub$Env), env_levels),
    dtype = torch_long()
  )

  n_envs <- length(env_levels)
  model  <- SyntheticVIModel(n_bands = length(BAND_NAMES),
                              n_envs = n_envs, d_embed = 4L, pred_h = 8L)
  opt    <- optim_adam(model$parameters, lr = lr, weight_decay = 1e-4)
  sched  <- lr_step(opt, step_size = 100, gamma = 0.5)

  model$train()
  for (ep in seq_len(epochs)) {
    idx <- sample(n)
    for (i in seq(1, n, by = batch_size)) {
      bi <- idx[i:min(i + batch_size - 1, n)]
      opt$zero_grad()
      pred <- model(tensors$bands[bi, , drop = FALSE],
                    env_idx = env_idx[bi])
      loss <- nnf_mse_loss(pred, tensors$y[bi, , drop = FALSE])
      loss$backward()
      nn_utils_clip_grad_norm_(model$parameters, 1.0)
      opt$step()
    }
    sched$step()
    if (verbose && ep %% 100 == 0)
      cat(sprintf("  ep %d | loss %.4f\n", ep, loss$item()))
  }
  model$eval()
  model
}

# ── LOEO: Leave-One-Environment-Out CV ────────────────────────────────────
# Takes raw (unscaled) data. Scaler fit on training environments only.
# For the unseen held-out environment, the mean embedding of the 4 training
# environments is used as a proxy in the prediction head.
loeo_cv <- function(data_raw, trait, env_levels, n_reps = 3,
                    epochs = 300, lr = 3e-3, batch_size = 64) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(held_env) {
    train_raw <- data_raw %>% filter(Env != held_env, !is.na(.data[[trait]]))
    test_raw  <- data_raw %>% filter(Env == held_env,  !is.na(.data[[trait]]))
    if (nrow(train_raw) < 10 || nrow(test_raw) < 2) return(NULL)

    sc_out   <- scale_from_train(train_raw, train_raw, trait)
    train_sc <- sc_out$df
    trait_mn <- sc_out$trait_mn
    trait_sd <- sc_out$trait_sd
    test_sc  <- scale_from_train(test_raw, train_raw, trait)$df

    # Indices (1-based) of the 4 training environments
    train_env_idx <- torch_tensor(
      match(sort(unique(as.character(train_raw$Env))), env_levels),
      dtype = torch_long()
    )

    map_dfr(seq_len(n_reps), function(rep_i) {
      model <- tryCatch(
        train_model(train_sc, trait, env_levels = env_levels,
                    epochs = epochs, lr = lr,
                    batch_size = batch_size, seed = rep_i * 100L),
        error = function(e) { message(held_env, " rep", rep_i, ": ", e$message); NULL }
      )
      if (is.null(model)) return(NULL)

      # Mean embedding of training envs → proxy for unseen held-out environment
      mean_env_vec <- with_no_grad({
        model$env_emb(train_env_idx)$mean(dim = 1, keepdim = TRUE)
      })

      te_ten <- make_tensors(test_sc, trait)
      with_no_grad({
        pred_z <- model(te_ten$bands, env_vec = mean_env_vec)$squeeze()$to(device = "cpu")
      })
      pred <- as.numeric(pred_z) * trait_sd + trait_mn
      obs  <- as.numeric(te_ten$y$squeeze()) * trait_sd + trait_mn
      tibble(held_env = held_env, rep = rep_i, n_test = length(obs),
             r    = cor(pred, obs, use = "complete.obs"),
             rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
    })
  })
}

# ── Within-environment 5-fold CV (split by GID) ───────────────────────────
# Takes raw (unscaled) data. GIDs assigned to folds; scaler fit on train GIDs.
# The correct environment index is known for both train and test, so the
# learned embedding is used directly (no mean-proxy needed).
within_env_cv <- function(data_raw, trait, env_levels, n_folds = 5, n_reps = 3,
                          epochs = 300, lr = 3e-3, batch_size = 64) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(env_i) {
    env_df    <- data_raw %>% filter(Env == env_i, !is.na(.data[[trait]]))
    gids      <- unique(env_df$GID)
    env_i_idx <- match(env_i, env_levels)   # 1-based index for this env (torch R convention)
    if (length(gids) < n_folds) {
      message("Env ", env_i, ": ", length(gids), " GIDs < n_folds — skipping")
      return(NULL)
    }
    map_dfr(seq_len(n_reps), function(rep_i) {
      set.seed(rep_i * 77L)
      shuffled <- sample(gids)
      fold_id  <- rep(seq_len(n_folds), length.out = length(shuffled))
      gid_fold <- tibble(GID = shuffled, fold = fold_id)

      map_dfr(seq_len(n_folds), function(f) {
        test_gids   <- gid_fold$GID[gid_fold$fold == f]
        train_raw_f <- env_df %>% filter(!GID %in% test_gids)
        test_raw_f  <- env_df %>% filter( GID %in% test_gids)
        if (nrow(train_raw_f) < 10 || nrow(test_raw_f) < 2) return(NULL)

        sc_out   <- scale_from_train(train_raw_f, train_raw_f, trait)
        train_sc <- sc_out$df
        trait_mn <- sc_out$trait_mn
        trait_sd <- sc_out$trait_sd
        test_sc  <- scale_from_train(test_raw_f, train_raw_f, trait)$df

        model <- tryCatch(
          train_model(train_sc, trait, env_levels = env_levels,
                      epochs = epochs, lr = lr,
                      batch_size = batch_size, seed = rep_i * 100L + f),
          error = function(e) NULL
        )
        if (is.null(model)) return(NULL)
        te_ten    <- make_tensors(test_sc, trait)
        te_env_idx <- torch_tensor(
          rep(env_i_idx, te_ten$bands$size(1)),
          dtype = torch_long()
        )
        with_no_grad({
          pred_z <- model(te_ten$bands,
                          env_idx = te_env_idx)$squeeze()$to(device = "cpu")
        })
        pred <- as.numeric(pred_z) * trait_sd + trait_mn
        obs  <- as.numeric(te_ten$y$squeeze()) * trait_sd + trait_mn
        tibble(Env = env_i, fold = f, rep = rep_i, n_test = length(obs),
               r    = cor(pred, obs, use = "complete.obs"),
               rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
      })
    })
  })
}

# ── Linear-model LOEO CV (for NDVI baseline) ─────────────────────────────
linreg_loeo_cv <- function(data_raw, predictor, trait) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(held_env) {
    tr <- data_raw %>% filter(Env != held_env,
                              !is.na(.data[[predictor]]), !is.na(.data[[trait]]))
    te <- data_raw %>% filter(Env == held_env,
                              !is.na(.data[[predictor]]), !is.na(.data[[trait]]))
    if (nrow(tr) < 10 || nrow(te) < 2) return(NULL)
    fit  <- lm(as.formula(paste(trait, "~", predictor)), data = tr)
    pred <- predict(fit, newdata = te)
    obs  <- te[[trait]]
    tibble(held_env = held_env, n_test = nrow(te),
           r    = cor(pred, obs, use = "complete.obs"),
           rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
  })
}

# ── Linear-model within-env 5-fold CV by GID (for NDVI baseline) ──────────
linreg_wenv_cv <- function(data_raw, predictor, trait,
                           n_folds = 5, n_reps = 3) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(env_i) {
    env_df <- data_raw %>% filter(Env == env_i,
                                  !is.na(.data[[predictor]]), !is.na(.data[[trait]]))
    gids <- unique(env_df$GID)
    if (length(gids) < n_folds) return(NULL)
    map_dfr(seq_len(n_reps), function(rep_i) {
      set.seed(rep_i * 77L)
      shuffled <- sample(gids)
      fold_id  <- rep(seq_len(n_folds), length.out = length(shuffled))
      gid_fold <- tibble(GID = shuffled, fold = fold_id)
      map_dfr(seq_len(n_folds), function(f) {
        test_gids <- gid_fold$GID[gid_fold$fold == f]
        tr <- env_df %>% filter(!GID %in% test_gids)
        te <- env_df %>% filter( GID %in% test_gids)
        if (nrow(tr) < 5 || nrow(te) < 2) return(NULL)
        fit  <- lm(as.formula(paste(trait, "~", predictor)), data = tr)
        pred <- predict(fit, newdata = te)
        obs  <- te[[trait]]
        tibble(Env = env_i, fold = f, rep = rep_i, n_test = nrow(te),
               r    = cor(pred, obs, use = "complete.obs"),
               rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
      })
    })
  })
}

# ── Helper: extract Synthetic VI values from a frozen encoder ─────────────
extract_synvi <- function(model, df) {
  if (nrow(df) == 0) return(numeric(0))
  bands_t <- torch_tensor(as.matrix(df[, BAND_NAMES, drop = FALSE]),
                          dtype = torch_float())
  with_no_grad({
    vi <- model$get_synthetic_vi(bands_t)$squeeze()
    if (vi$dim() == 0L) vi <- vi$unsqueeze(1L)
    as.numeric(vi$to(device = "cpu"))
  })
}

# ── Variant B: pre-train on 4 envs → freeze encoder → fine-tune head ───────
# Pre-trains on the other 4 environments, then for each fold within the target
# environment freezes the encoder and fine-tunes only the prediction head and
# environment embeddings on the adaptation (train) GIDs, then predicts the test GIDs.
adapt_finetune_cv <- function(data_raw, trait, env_levels,
                               n_folds        = 5,   n_reps         = 3,
                               pretrain_epochs = 300, pretrain_lr    = 3e-3,
                               finetune_epochs = 50,  finetune_lr    = 1e-3,
                               batch_size      = 64) {
  envs <- sort(unique(as.character(data_raw$Env)))

  map_dfr(envs, function(target_env) {
    train_4env <- data_raw %>% filter(Env != target_env, !is.na(.data[[trait]]))
    target_df  <- data_raw %>% filter(Env == target_env,  !is.na(.data[[trait]]))
    gids       <- unique(target_df$GID)
    if (nrow(train_4env) < 10 || length(gids) < n_folds) return(NULL)

    # Scaler fit on 4 training environments — applied to adaptation & test folds too
    sc_out   <- scale_from_train(train_4env, train_4env, trait)
    train_sc <- sc_out$df
    trait_mn <- sc_out$trait_mn
    trait_sd <- sc_out$trait_sd

    target_env_i <- match(target_env, env_levels)  # 1-based embedding index

    map_dfr(seq_len(n_reps), function(rep_i) {
      # Step 1: pre-train on 4 environments
      pretrained <- tryCatch(
        train_model(train_sc, trait, env_levels = env_levels,
                    epochs = pretrain_epochs, lr = pretrain_lr,
                    batch_size = batch_size, seed = rep_i * 100L),
        error = function(e) { message(target_env, " rep", rep_i, ": ", e$message); NULL }
      )
      if (is.null(pretrained)) return(NULL)

      pretrained_state <- pretrained$state_dict()  # snapshot before any fine-tuning

      # K-fold CV within target environment (by GID)
      set.seed(rep_i * 77L)
      shuffled <- sample(gids)
      fold_id  <- rep(seq_len(n_folds), length.out = length(shuffled))
      gid_fold <- tibble(GID = shuffled, fold = fold_id)

      map_dfr(seq_len(n_folds), function(f) {
        test_gids  <- gid_fold$GID[gid_fold$fold == f]
        adapt_gids <- gids[!gids %in% test_gids]
        adapt_df   <- target_df %>% filter(GID %in% adapt_gids)
        test_df    <- target_df %>% filter(GID %in% test_gids)
        if (nrow(adapt_df) < 5 || nrow(test_df) < 2) return(NULL)

        # Scale with train_4env stats (bands 0-1, trait z-scored for head loss)
        adapt_sc <- scale_from_train(adapt_df, train_4env, trait)$df
        test_sc  <- scale_from_train(test_df,  train_4env, trait)$df

        # Fresh model initialised from pre-trained weights (clean copy per fold)
        model_ft <- SyntheticVIModel(n_bands = length(BAND_NAMES),
                                      n_envs  = length(env_levels),
                                      d_embed = 4L, pred_h = 8L)
        model_ft$load_state_dict(pretrained_state)

        # Freeze encoder — gradients disabled for all BandEncoder parameters
        for (p in model_ft$encoder$parameters) p$requires_grad_(FALSE)

        # Collect only trainable parameters (head + embeddings)
        trainable <- list(
          model_ft$env_emb$weight,
          model_ft$pred_fc1$weight, model_ft$pred_fc1$bias,
          model_ft$pred_fc2$weight, model_ft$pred_fc2$bias
        )
        opt_ft <- optim_adam(trainable, lr = finetune_lr, weight_decay = 1e-4)

        adapt_ten     <- make_tensors(adapt_sc, trait)
        n_adapt       <- adapt_ten$bands$size(1)
        adapt_env_idx <- torch_tensor(rep(target_env_i, n_adapt), dtype = torch_long())

        # Step 2: fine-tune head on adaptation fold
        model_ft$train()
        for (ep in seq_len(finetune_epochs)) {
          idx_s <- sample(n_adapt)
          for (i in seq(1, n_adapt, by = batch_size)) {
            bi <- idx_s[i:min(i + batch_size - 1, n_adapt)]
            # Recreate env index tensor directly to avoid 0D scalar from single-element indexing
            ei <- torch_tensor(rep(target_env_i, length(bi)), dtype = torch_long())
            opt_ft$zero_grad()
            pred_ft <- model_ft(adapt_ten$bands[bi, , drop = FALSE], env_idx = ei)
            loss <- nnf_mse_loss(pred_ft, adapt_ten$y[bi, , drop = FALSE])
            loss$backward()
            nn_utils_clip_grad_norm_(trainable, 1.0)
            opt_ft$step()
          }
        }
        model_ft$eval()

        # Step 3: predict test fold
        test_ten   <- make_tensors(test_sc, trait)
        te_env_idx <- torch_tensor(
          rep(target_env_i, test_ten$bands$size(1)), dtype = torch_long()
        )
        with_no_grad({
          pred_z <- model_ft(test_ten$bands, env_idx = te_env_idx)$squeeze()$to(device = "cpu")
        })
        pred <- as.numeric(pred_z) * trait_sd + trait_mn
        obs  <- as.numeric(test_ten$y$squeeze()) * trait_sd + trait_mn

        tibble(Env = target_env, fold = f, rep = rep_i, n_test = length(obs),
               r    = cor(pred, obs, use = "complete.obs"),
               rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
      })
    })
  })
}

# ── Variant C: pre-train on 4 envs → freeze encoder → linear regression ────
# Pre-trains on the other 4 environments, extracts the Synthetic VI from the
# frozen encoder, then fits a simple OLS (trait ~ SynVI) on the adaptation GIDs
# and predicts the test GIDs. No gradient updates to the DNN in this stage.
adapt_lm_cv <- function(data_raw, trait, env_levels,
                         n_folds        = 5, n_reps      = 3,
                         pretrain_epochs = 300, pretrain_lr = 3e-3,
                         batch_size      = 64) {
  envs <- sort(unique(as.character(data_raw$Env)))

  map_dfr(envs, function(target_env) {
    train_4env <- data_raw %>% filter(Env != target_env, !is.na(.data[[trait]]))
    target_df  <- data_raw %>% filter(Env == target_env,  !is.na(.data[[trait]]))
    gids       <- unique(target_df$GID)
    if (nrow(train_4env) < 10 || length(gids) < n_folds) return(NULL)

    sc_out   <- scale_from_train(train_4env, train_4env, trait)
    train_sc <- sc_out$df

    map_dfr(seq_len(n_reps), function(rep_i) {
      # Step 1: pre-train encoder on 4 environments
      pretrained <- tryCatch(
        train_model(train_sc, trait, env_levels = env_levels,
                    epochs = pretrain_epochs, lr = pretrain_lr,
                    batch_size = batch_size, seed = rep_i * 100L),
        error = function(e) { message(target_env, " rep", rep_i, ": ", e$message); NULL }
      )
      if (is.null(pretrained)) return(NULL)
      pretrained$eval()

      # K-fold CV within target environment (by GID)
      set.seed(rep_i * 77L)
      shuffled <- sample(gids)
      fold_id  <- rep(seq_len(n_folds), length.out = length(shuffled))
      gid_fold <- tibble(GID = shuffled, fold = fold_id)

      map_dfr(seq_len(n_folds), function(f) {
        test_gids  <- gid_fold$GID[gid_fold$fold == f]
        adapt_gids <- gids[!gids %in% test_gids]
        adapt_df   <- target_df %>% filter(GID %in% adapt_gids)
        test_df    <- target_df %>% filter(GID %in% test_gids)
        if (nrow(adapt_df) < 5 || nrow(test_df) < 2) return(NULL)

        # Scale bands using train_4env scaler (encoder trained on this scale)
        adapt_sc_bands <- scale_from_train(adapt_df, train_4env, trait)$df
        test_sc_bands  <- scale_from_train(test_df,  train_4env, trait)$df

        # Step 2: extract SynVI from frozen encoder (no gradient updates)
        synvi_adapt <- extract_synvi(pretrained, adapt_sc_bands)
        synvi_test  <- extract_synvi(pretrained, test_sc_bands)

        # Step 3: fit OLS on original-scale trait ~ SynVI, predict test
        adapt_y <- adapt_df[[trait]]
        test_y  <- test_df[[trait]]

        lm_fit <- lm(y ~ synvi, data.frame(y = adapt_y, synvi = synvi_adapt))
        pred   <- predict(lm_fit, data.frame(synvi = synvi_test))

        tibble(Env = target_env, fold = f, rep = rep_i, n_test = length(test_y),
               r    = cor(pred, test_y, use = "complete.obs"),
               rmse = sqrt(mean((pred - test_y)^2, na.rm = TRUE)))
      })
    })
  })
}
```

```{r run-training, cache=TRUE}
# ── Select traits with enough observations ────────────────────────────────
min_obs <- 30
usable_traits <- trait_cols[
  sapply(trait_cols, function(t) sum(!is.na(model_data_raw[[t]]))) >= min_obs
]
cat("Traits with >=", min_obs, "observations:", length(usable_traits), "\n")
print(usable_traits)

usable_traits <- usable_traits[1]   # user selection – expand as needed

# ── Environment levels: consistent ordering used for embedding indices ────
env_levels <- sort(unique(as.character(model_data_raw$Env)))
cat("Environment levels (embedding order):", paste(env_levels, collapse = ", "), "\n")

# ── LOEO CV (raw data → per-fold scaling inside loeo_cv) ─────────────────
cat("\nRunning LOEO CV...\n")
cv_loeo <- map_dfr(usable_traits, function(trait) {
  cat("  trait:", trait, "\n")
  res       <- loeo_cv(model_data_raw, trait, env_levels = env_levels,
                       n_reps = 3, epochs = 300, lr = 3e-3, batch_size = 64)
  res$trait <- trait
  res
})
cat("LOEO CV done. Rows:", nrow(cv_loeo), "\n")

# ── Within-environment 5-fold CV (raw data → per-fold scaling inside) ────
cat("\nRunning within-environment 5-fold CV...\n")
cv_wenv <- map_dfr(usable_traits, function(trait) {
  cat("  trait:", trait, "\n")
  res       <- within_env_cv(model_data_raw, trait, env_levels = env_levels,
                              n_folds = 5, n_reps = 3,
                              epochs = 300, lr = 3e-3, batch_size = 64)
  res$trait <- trait
  res
})
cat("Within-env CV done. Rows:", nrow(cv_wenv), "\n")
```

```{r ndvi-baseline}
# ── NDVI linear-regression baseline (same CV design as DNN) ─────────────
if (!"NDVI_raw" %in% names(model_data_raw)) {
  warning("NDVI_raw not found in model_data_raw — skipping NDVI baseline.")
  ndvi_loeo <- tibble()
  ndvi_wenv <- tibble()
} else {
  cat("Running NDVI LOEO baseline CV...\n")
  ndvi_loeo <- map_dfr(usable_traits, function(trait) {
    res       <- linreg_loeo_cv(model_data_raw, "NDVI_raw", trait)
    res$trait <- trait
    res
  })
  cat("NDVI LOEO done. Rows:", nrow(ndvi_loeo), "\n")

  cat("Running NDVI within-env baseline CV...\n")
  ndvi_wenv <- map_dfr(usable_traits, function(trait) {
    res       <- linreg_wenv_cv(model_data_raw, "NDVI_raw", trait)
    res$trait <- trait
    res
  })
  cat("NDVI within-env done. Rows:", nrow(ndvi_wenv), "\n")
}
```

```{r center-comparison}
# ── Within-environment mean-centering comparison ───────────────────────────
# For each band and the trait: subtract the environment mean (not SD-scaled).
# NDVI is computed from original reflectances then centered by env mean NDVI.
# At LOEO test time each environment is centered on its own mean, so no
# information leaks between training and test environments.

center_within_env <- function(data_raw, trait) {
  data_raw %>%
    group_by(Env) %>%
    mutate(
      across(all_of(BAND_NAMES), ~ . - mean(., na.rm = TRUE)),
      !!trait        := .data[[trait]] - mean(.data[[trait]], na.rm = TRUE),
      NDVI_centered  = NDVI_raw - mean(NDVI_raw, na.rm = TRUE)
    ) %>%
    ungroup()
}

trait <- usable_traits[1]

# Build centered dataset from original (un-z-scored) reflectances
model_data_centered <- center_within_env(
  model_data_raw_orig %>% filter(Env != "MCG25"), trait
)

cat("Running DNN LOEO CV on centered data...\n")
cv_loeo_centered <- map_dfr(usable_traits, function(tr) {
  res       <- loeo_cv(model_data_centered, tr, env_levels = env_levels,
                       n_reps = 3, epochs = 300, lr = 3e-3, batch_size = 64)
  res$trait <- tr
  res
})
cat("DNN centered LOEO done. Rows:", nrow(cv_loeo_centered), "\n")

cat("Running NDVI LOEO CV on centered data...\n")
ndvi_loeo_centered <- map_dfr(usable_traits, function(tr) {
  res       <- linreg_loeo_cv(model_data_centered, "NDVI_centered", tr)
  res$trait <- tr
  res
})
cat("NDVI centered LOEO done. Rows:", nrow(ndvi_loeo_centered), "\n")
```

# Results: Cross-Validation Accuracy

## Leave-One-Environment-Out (LOEO)

```{r loeo-summary-table}
loeo_fold <- cv_loeo %>%
  filter(!is.na(r)) %>%
  group_by(trait, held_env) %>%
  summarise(r      = mean(r,      na.rm = TRUE),
            rmse   = mean(rmse,   na.rm = TRUE),
            n_test = first(n_test),
            .groups = "drop")

loeo_summary <- loeo_fold %>%
  group_by(trait) %>%
  summarise(mean_r    = mean(r,    na.rm = TRUE),
            sd_r      = sd(r,     na.rm = TRUE),
            mean_rmse = mean(rmse, na.rm = TRUE),
            n_folds   = n(),
            .groups   = "drop") %>%
  arrange(desc(mean_r))

loeo_summary %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r","SD r","Mean RMSE","Folds"),
        caption   = "LOEO CV: DNN Synthetic VI accuracy (mean over 5 held-out environments, 3 training reps each)") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(loeo_summary$mean_r > 0.5, "green", "black"))

cat("\nPer-environment LOEO results:\n")
loeo_fold %>%
  arrange(trait, held_env) %>%
  kable(digits = 3,
        col.names = c("Trait","Held-out env","Mean r","Mean RMSE","N test"),
        caption   = "LOEO CV per held-out environment") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r loeo-accuracy-plot, fig.height=5, fig.cap="LOEO CV: dots = training repetitions, crossbar = mean r"}
cv_loeo %>%
  filter(!is.na(r)) %>%
  ggplot(aes(x = r, y = reorder(held_env, r, mean, na.rm = TRUE),
             colour = held_env)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(height = 0.12, size = 2.5, alpha = 0.7) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45,
               colour = "black", linewidth = 0.7) +
  facet_wrap(~trait, ncol = 2) +
  scale_colour_brewer(palette = "Set2") +
  labs(x = "Pearson r (LOEO)", y = "Held-out environment",
       title  = "DNN Synthetic VI – LOEO Accuracy",
       colour = "Environment") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

## Within-Environment 5-Fold CV (by GID)

```{r wenv-summary-table}
wenv_fold_summ <- cv_wenv %>%
  filter(!is.na(r)) %>%
  group_by(trait, Env, fold) %>%
  summarise(r    = mean(r,    na.rm = TRUE),
            rmse = mean(rmse, na.rm = TRUE),
            .groups = "drop")

wenv_env_summary <- wenv_fold_summ %>%
  group_by(trait, Env) %>%
  summarise(mean_r    = mean(r,    na.rm = TRUE),
            sd_r      = sd(r,     na.rm = TRUE),
            mean_rmse = mean(rmse, na.rm = TRUE),
            n_folds   = n(),
            .groups   = "drop")

wenv_summary <- wenv_env_summary %>%
  group_by(trait) %>%
  summarise(mean_r    = mean(mean_r,    na.rm = TRUE),
            sd_r      = mean(sd_r,      na.rm = TRUE),
            mean_rmse = mean(mean_rmse, na.rm = TRUE),
            .groups   = "drop")

wenv_summary %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r","Mean within-env SD r","Mean RMSE"),
        caption   = "Within-environment 5-fold CV (by GID): mean Pearson r across all environments and folds") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(wenv_summary$mean_r > 0.5, "green", "black"))

cat("\nPer-environment within-env CV results:\n")
wenv_env_summary %>%
  arrange(trait, Env) %>%
  kable(digits = 3,
        col.names = c("Trait","Environment","Mean r","SD r","Mean RMSE","N folds"),
        caption   = "Within-environment 5-fold CV: per-environment summary") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r wenv-accuracy-plot, fig.height=5, fig.cap="Within-env CV: dots = individual folds × reps, crossbar = mean r per environment"}
cv_wenv %>%
  filter(!is.na(r)) %>%
  ggplot(aes(x = r, y = reorder(Env, r, mean, na.rm = TRUE),
             colour = Env)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(height = 0.12, size = 2, alpha = 0.6) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45,
               colour = "black", linewidth = 0.7) +
  facet_wrap(~trait, ncol = 2) +
  scale_colour_brewer(palette = "Set2") +
  labs(x = "Pearson r (within-env 5-fold CV by GID)", y = "Environment",
       title  = "DNN Synthetic VI – Within-Environment CV Accuracy",
       colour = "Environment") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

## CV Scenario Comparison

```{r cv-scenario-comparison}
cv_scenario_tbl <- bind_rows(
  loeo_summary  %>% select(trait, mean_r, sd_r, mean_rmse) %>% mutate(CV = "LOEO"),
  wenv_summary  %>% select(trait, mean_r, sd_r, mean_rmse) %>% mutate(CV = "Within-env 5-fold")
)

cv_scenario_tbl %>%
  arrange(trait, CV) %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r","SD r","Mean RMSE","CV scenario"),
        caption   = "DNN Synthetic VI: LOEO vs. within-environment CV") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r cv-scenario-plot, fig.height=4, fig.cap="LOEO vs. within-environment CV: mean Pearson r (error bar = SD across folds/envs)"}
cv_scenario_tbl %>%
  ggplot(aes(x = mean_r,
             y = reorder(trait, mean_r, max, na.rm = TRUE),
             colour = CV, shape = CV)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = mean_r - sd_r, xmax = mean_r + sd_r),
                 height = 0.2, alpha = 0.6) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(LOEO = "#e34a33",
                                 `Within-env 5-fold` = "#2c7bb6")) +
  labs(x = "Mean Pearson r", y = NULL,
       title  = "LOEO vs. within-environment CV accuracy",
       colour = "CV scenario", shape = "CV scenario") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

# NDVI Baseline Comparison

NDVI-based linear regression was run under the same two CV schemes (LOEO and
within-environment 5-fold CV by GID) as the DNN, operating on raw (unscaled)
trait values so RMSE is directly comparable.

## NDVI Baseline — LOEO

```{r ndvi-loeo-table}
if (nrow(ndvi_loeo) > 0) {
  ndvi_loeo_fold <- ndvi_loeo %>%
    filter(!is.na(r)) %>%
    group_by(trait, held_env) %>%
    summarise(r    = mean(r,    na.rm = TRUE),
              rmse = mean(rmse, na.rm = TRUE),
              .groups = "drop")

  ndvi_loeo_summary <- ndvi_loeo_fold %>%
    group_by(trait) %>%
    summarise(ndvi_r    = mean(r,    na.rm = TRUE),
              ndvi_rmse = mean(rmse, na.rm = TRUE),
              .groups = "drop")

  ndvi_loeo_summary %>%
    kable(digits = 3,
          col.names = c("Trait", "NDVI mean r (LOEO)", "NDVI mean RMSE"),
          caption   = "NDVI linear regression — LOEO accuracy") %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE)
} else {
  cat("NDVI baseline not available.\n")
}
```

## NDVI Baseline — Within-Environment 5-Fold CV

```{r ndvi-wenv-table}
if (nrow(ndvi_wenv) > 0) {
  ndvi_wenv_fold <- ndvi_wenv %>%
    filter(!is.na(r)) %>%
    group_by(trait, Env, fold) %>%
    summarise(r    = mean(r,    na.rm = TRUE),
              rmse = mean(rmse, na.rm = TRUE),
              .groups = "drop")

  ndvi_wenv_summary <- ndvi_wenv_fold %>%
    group_by(trait, Env) %>%
    summarise(mean_r    = mean(r,    na.rm = TRUE),
              mean_rmse = mean(rmse, na.rm = TRUE),
              .groups   = "drop") %>%
    group_by(trait) %>%
    summarise(ndvi_r    = mean(mean_r,    na.rm = TRUE),
              ndvi_rmse = mean(mean_rmse, na.rm = TRUE),
              .groups   = "drop")

  ndvi_wenv_summary %>%
    kable(digits = 3,
          col.names = c("Trait", "NDVI mean r (within-env)", "NDVI mean RMSE"),
          caption   = "NDVI linear regression — within-environment 5-fold CV accuracy") %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE)
} else {
  cat("NDVI baseline not available.\n")
}
```

## DNN Synthetic VI vs. NDVI — Combined Comparison

```{r ndvi-dnn-comparison-table}
if (nrow(ndvi_loeo) > 0 && nrow(ndvi_wenv) > 0) {
  dnn_loeo_r  <- loeo_summary  %>% select(trait, dnn_loeo_r  = mean_r)
  dnn_wenv_r  <- wenv_summary  %>% select(trait, dnn_wenv_r  = mean_r)
  ndvi_loeo_r <- ndvi_loeo_summary %>% select(trait, ndvi_loeo_r = ndvi_r)
  ndvi_wenv_r <- ndvi_wenv_summary %>% select(trait, ndvi_wenv_r = ndvi_r)

  full_comparison <- dnn_loeo_r %>%
    left_join(ndvi_loeo_r, by = "trait") %>%
    left_join(dnn_wenv_r,  by = "trait") %>%
    left_join(ndvi_wenv_r, by = "trait") %>%
    mutate(delta_loeo = dnn_loeo_r - ndvi_loeo_r,
           delta_wenv = dnn_wenv_r - ndvi_wenv_r)

  full_comparison %>%
    arrange(desc(delta_loeo)) %>%
    kable(digits = 3,
          col.names = c("Trait",
                        "DNN r (LOEO)", "NDVI r (LOEO)", "Δr LOEO",
                        "DNN r (within-env)", "NDVI r (within-env)", "Δr within-env"),
          caption = "DNN Synthetic VI vs. NDVI linear regression — both CV scenarios") %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE) %>%
    column_spec(4, bold = TRUE,
                color = ifelse(full_comparison$delta_loeo > 0, "forestgreen", "firebrick")) %>%
    column_spec(7, bold = TRUE,
                color = ifelse(full_comparison$delta_wenv > 0, "forestgreen", "firebrick"))
}
```

```{r ndvi-dnn-comparison-plot, fig.height=5, fig.cap="DNN Synthetic VI vs. NDVI linear regression — Pearson r by CV scenario"}
if (nrow(ndvi_loeo) > 0 && nrow(ndvi_wenv) > 0 && exists("full_comparison")) {
  plot_df <- bind_rows(
    full_comparison %>%
      select(trait, r = dnn_loeo_r)  %>% mutate(model = "DNN Synthetic VI",    CV = "LOEO"),
    full_comparison %>%
      select(trait, r = ndvi_loeo_r) %>% mutate(model = "NDVI (linear)",       CV = "LOEO"),
    full_comparison %>%
      select(trait, r = dnn_wenv_r)  %>% mutate(model = "DNN Synthetic VI",    CV = "Within-env 5-fold"),
    full_comparison %>%
      select(trait, r = ndvi_wenv_r) %>% mutate(model = "NDVI (linear)",       CV = "Within-env 5-fold")
  )

  ggplot(plot_df,
         aes(x = r, y = reorder(trait, r, max, na.rm = TRUE),
             colour = model, shape = model)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_point(size = 4, alpha = 0.9) +
    geom_line(aes(group = interaction(trait, CV)), colour = "grey60",
              linewidth = 0.5) +
    facet_wrap(~CV, ncol = 2) +
    scale_colour_manual(values = c("DNN Synthetic VI" = "#e34a33",
                                   "NDVI (linear)"    = "#3182bd")) +
    scale_shape_manual(values  = c("DNN Synthetic VI" = 16,
                                   "NDVI (linear)"    = 17)) +
    labs(x = "Mean Pearson r", y = NULL,
         title  = "DNN Synthetic VI vs. NDVI baseline",
         colour = NULL, shape = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

## Within-Environment Centering: 4-Way Comparison

Compares LOEO accuracy across four modelling strategies: DNN and NDVI linear
regression, each run on **uncentered** (per-env z-scored bands) and
**within-centered** (env mean subtracted from bands and trait) data.

```{r centering-comparison-table}
# Summarise each of the four LOEO variants per environment
summarise_loeo <- function(cv_df, label) {
  cv_df %>%
    filter(!is.na(r)) %>%
    group_by(held_env) %>%
    summarise(r = mean(r, na.rm = TRUE), .groups = "drop") %>%
    mutate(model = label)
}

loeo_4way <- bind_rows(
  summarise_loeo(cv_loeo,           "DNN (z-scored)"),
  summarise_loeo(cv_loeo_centered,  "DNN (centered)"),
  summarise_loeo(ndvi_loeo %>% rename(held_env = held_env),
                 "NDVI linear (uncentered)"),
  summarise_loeo(ndvi_loeo_centered %>% rename(held_env = held_env),
                 "NDVI linear (centered)")
)

loeo_4way_wide <- loeo_4way %>%
  pivot_wider(names_from = model, values_from = r) %>%
  rename(Environment = held_env) %>%
  bind_rows(
    summarise(., across(where(is.numeric), ~ round(mean(., na.rm = TRUE), 3)),
              Environment = "**Mean**")
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

loeo_4way_wide %>%
  kable(digits = 3,
        caption = "LOEO Pearson r — 4-way comparison: DNN vs. NDVI linear × uncentered vs. within-centered") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r centering-comparison-plot, fig.height=5, fig.cap="LOEO Pearson r by environment — 4 model variants"}
loeo_4way %>%
  mutate(model = factor(model,
                        levels = c("DNN (z-scored)", "DNN (centered)",
                                   "NDVI linear (uncentered)", "NDVI linear (centered)"))) %>%
  ggplot(aes(x = r, y = reorder(held_env, r, mean, na.rm = TRUE),
             colour = model, shape = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 4, alpha = 0.9) +
  scale_colour_manual(values = c("DNN (z-scored)"          = "#e34a33",
                                 "DNN (centered)"           = "#fc8d59",
                                 "NDVI linear (uncentered)" = "#2166ac",
                                 "NDVI linear (centered)"   = "#74add1")) +
  scale_shape_manual(values  = c("DNN (z-scored)"          = 16,
                                 "DNN (centered)"           = 17,
                                 "NDVI linear (uncentered)" = 15,
                                 "NDVI linear (centered)"   = 18)) +
  labs(x = "LOEO Pearson r", y = "Held-out environment",
       title  = "LOEO accuracy: DNN vs. NDVI × uncentered vs. within-centered",
       colour = NULL, shape = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

# Transfer Learning: Two-Stage Adaptation CV

Both variants share the same outer structure: **pre-train** the full model on the
four non-target environments, then perform K-fold CV **within** the target
environment using the pre-trained encoder.

- **Variant B (fine-tune head)**: the encoder is frozen; only the prediction head
  and environment embeddings are updated on the adaptation (train) GIDs.
  The target environment's embedding was never seen during pre-training, so it
  starts from its random initialisation and is calibrated during fine-tuning.

- **Variant C (SynVI + linear)**: the encoder is completely frozen; the Synthetic
  VI is extracted for every observation, and a simple OLS (trait ~ SynVI) is fit
  on the adaptation GIDs and used to predict the test GIDs — no further neural
  network training.

Both are compared to the standard within-environment 5-fold CV that trains from
scratch on each fold.

```{r run-adaptation-cv, cache=TRUE}
cat("Running Variant B: freeze encoder, fine-tune head...\n")
cv_adapt_finetune <- map_dfr(usable_traits, function(tr) {
  cat("  trait:", tr, "\n")
  res       <- adapt_finetune_cv(model_data_raw, tr, env_levels = env_levels,
                                  n_folds        = 5,   n_reps         = 3,
                                  pretrain_epochs = 300, pretrain_lr    = 3e-3,
                                  finetune_epochs = 50,  finetune_lr    = 1e-3,
                                  batch_size      = 64)
  res$trait <- tr
  res
})
cat("Variant B done. Rows:", nrow(cv_adapt_finetune), "\n")

cat("\nRunning Variant C: freeze encoder, linear regression on SynVI...\n")
cv_adapt_lm <- map_dfr(usable_traits, function(tr) {
  cat("  trait:", tr, "\n")
  res       <- adapt_lm_cv(model_data_raw, tr, env_levels = env_levels,
                            n_folds        = 5,   n_reps      = 3,
                            pretrain_epochs = 300, pretrain_lr = 3e-3,
                            batch_size      = 64)
  res$trait <- tr
  res
})
cat("Variant C done. Rows:", nrow(cv_adapt_lm), "\n")
```

```{r adapt-summary-table}
summarise_adapt_env <- function(cv_df, label) {
  cv_df %>%
    filter(!is.na(r)) %>%
    group_by(trait, Env) %>%
    summarise(r = mean(r, na.rm = TRUE), .groups = "drop") %>%
    mutate(model = label)
}

adapt_comparison <- bind_rows(
  cv_wenv %>%
    filter(!is.na(r)) %>%
    group_by(trait, Env) %>%
    summarise(r = mean(r, na.rm = TRUE), .groups = "drop") %>%
    mutate(model = "Within-env (train from scratch)"),
  summarise_adapt_env(cv_adapt_finetune, "Adapt B: fine-tune head"),
  summarise_adapt_env(cv_adapt_lm,       "Adapt C: SynVI + linear")
)

adapt_overall <- adapt_comparison %>%
  group_by(trait, model) %>%
  summarise(mean_r = round(mean(r, na.rm = TRUE), 3),
            sd_r   = round(sd(r,   na.rm = TRUE), 3),
            .groups = "drop") %>%
  arrange(trait, model)

adapt_overall %>%
  kable(digits = 3,
        col.names = c("Trait", "Model", "Mean r", "SD r (across envs)"),
        caption = "Two-stage adaptation CV: mean within-environment Pearson r") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)

cat("\nPer-environment breakdown:\n")
adapt_comparison %>%
  pivot_wider(names_from = model, values_from = r) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  arrange(trait, Env) %>%
  kable(digits = 3,
        caption = "Pearson r per environment — three within-env CV strategies") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r adapt-comparison-plot, fig.height=5, fig.cap="Two-stage adaptation CV vs. standard within-env CV: Pearson r per environment"}
adapt_comparison %>%
  mutate(model = factor(model,
                        levels = c("Within-env (train from scratch)",
                                   "Adapt B: fine-tune head",
                                   "Adapt C: SynVI + linear"))) %>%
  ggplot(aes(x = r, y = reorder(Env, r, mean, na.rm = TRUE),
             colour = model, shape = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 4, alpha = 0.9) +
  facet_wrap(~trait, ncol = 2) +
  scale_colour_manual(values = c(
    "Within-env (train from scratch)" = "#2166ac",
    "Adapt B: fine-tune head"         = "#e34a33",
    "Adapt C: SynVI + linear"         = "#1a9850"
  )) +
  scale_shape_manual(values = c(
    "Within-env (train from scratch)" = 16,
    "Adapt B: fine-tune head"         = 17,
    "Adapt C: SynVI + linear"         = 15
  )) +
  labs(x = "Pearson r (within-env K-fold CV)", y = "Target environment",
       title  = "Two-stage adaptation CV vs. standard within-env training",
       colour = NULL, shape = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

# Full-Data Models and Synthetic VI Extraction

One final model per trait is trained on **all five groups** (no held-out fold).
These models are used to (a) extract the Synthetic VI for every observation and
(b) derive the VI formula from the learned encoder weights.

```{r extract-synthetic-vi}
full_models <- map(usable_traits, function(trait) {
  cat("Training full model – trait:", trait, "\n")
  tryCatch(
    train_model(model_data, trait, env_levels = env_levels,
                epochs = 300, lr = 3e-3, batch_size = 64, seed = 42L),
    error = function(e) { message(e$message); NULL }
  )
}) %>% set_names(usable_traits)

synvi_rows <- map_dfr(usable_traits, function(trait) {
  model <- full_models[[trait]]
  if (is.null(model)) return(NULL)
  df <- model_data %>% filter(!is.na(.data[[trait]]))
  bands_t <- torch_tensor(as.matrix(df[, BAND_NAMES, drop = FALSE]),
                          dtype = torch_float())
  with_no_grad({
    vi_vals <- model$get_synthetic_vi(bands_t)$squeeze()$to(device = "cpu")
  })
  df %>%
    select(Env, Timepoint, group) %>%
    mutate(SynVI = as.numeric(vi_vals), trait = trait)
})

cat("Synthetic VI extraction complete. Rows:", nrow(synvi_rows), "\n")
```

```{r synthetic-vi-distributions, fig.height=6, fig.cap="Synthetic VI distributions by Timepoint and Environment, per trait"}
synvi_rows %>%
  ggplot(aes(x = SynVI, fill = Env)) +
  geom_density(alpha = 0.5) +
  facet_grid(trait ~ Timepoint, scales = "free") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Synthetic VI value", y = "Density",
       title = "Synthetic VI distributions by Timepoint × Environment",
       fill  = "Environment") +
  theme_bw(base_size = 10) +
  theme(strip.text.y = element_text(size = 7))
```

# Band Sensitivity Analysis

To interpret *which* raw bands drive each Synthetic VI we compute the gradient
of the encoder output with respect to each band input (averaged over all samples).

```{r sensitivity-analysis}
# Use the full-data models for gradient-based band importance
sensitivity_list <- map_dfr(usable_traits, function(trait) {
  model <- full_models[[trait]]
  if (is.null(model)) return(NULL)
  df <- model_data %>% filter(!is.na(.data[[trait]]))
  bands_t <- torch_tensor(
    as.matrix(df[, BAND_NAMES, drop = FALSE]),
    dtype = torch_float(), requires_grad = TRUE
  )
  vi <- model$encoder(bands_t)
  vi$sum()$backward()
  grads    <- as.matrix(bands_t$grad$to(device = "cpu"))
  abs_mean <- colMeans(abs(grads))
  tibble(trait = trait,
         band  = BAND_NAMES,
         importance = abs_mean / sum(abs_mean))
})

cat("Sensitivity analysis complete.\n")
```

```{r sensitivity-heatmap, fig.cap="Normalised band importance for each Synthetic VI (gradient-based)"}
if (!is.null(sensitivity_list) && nrow(sensitivity_list) > 0) {
  imp_wide <- sensitivity_list %>%
    pivot_wider(names_from = band, values_from = importance)

  imp_mat <- as.matrix(imp_wide[, BAND_NAMES])
  rownames(imp_mat) <- imp_wide$trait

    sensitivity_list %>%
      ggplot(aes(x = band, y = reorder(trait, importance),
                 fill = importance)) +
      geom_tile(colour = "white") +
      scale_fill_gradient(low = "white", high = "#2166AC") +
      labs(x = "Spectral Band", y = "Trait",
           fill  = "Importance",
           title = "Relative Band Importance for each Synthetic VI") +
      theme_bw(base_size = 11)
  }

```

```{r sensitivity-barplot, fig.cap="Band importance profiles per trait"}
if (!is.null(sensitivity_list) && nrow(sensitivity_list) > 0) {
  ggplot(sensitivity_list,
         aes(x = band, y = importance, fill = band)) +
    geom_col() +
    facet_wrap(~trait, ncol = 3) +
    scale_fill_manual(values = c(blue="steelblue", green="#2ca25f",
                                 red="#e34a33", red_edge="#f768a1",
                                 nir="#8856a7")) +
    labs(x = "Band", y = "Normalised gradient importance",
         title = "Per-trait spectral band importance") +
    theme_bw(base_size = 11) +
    theme(axis.text.x  = element_text(angle = 30, hjust = 1),
          legend.position = "none")
}
```

# Synthetic VI Formula

The Synthetic VI is defined by the three learned linear layers of the encoder:

$$
\mathbf{h}_1 = \text{ReLU}(\mathbf{W}_1 \mathbf{x} + \mathbf{b}_1), \quad
\mathbf{h}_2 = \text{ReLU}(\mathbf{W}_2 \mathbf{h}_1 + \mathbf{b}_2), \quad
\text{SynVI} = \mathbf{w}_3^\top \mathbf{h}_2 + b_3
$$

where $\mathbf{x} = [\text{blue, green, red, red\_edge, nir}]^\top$ are the
min-max scaled band values. The tables below show the extracted weights for each
trait's full-data model.

```{r vi-formula, results="asis"}
for (trait in usable_traits) {
  model <- full_models[[trait]]
  if (is.null(model)) next

  cat("\n\n## Trait:", trait, "\n\n")

  # ── Extract layer weights ──────────────────────────────────────────────
  W1 <- as.matrix(model$encoder$fc1$weight$to(device="cpu")$detach())
  b1 <- as.numeric(model$encoder$fc1$bias$to(device="cpu")$detach())
  W2 <- as.matrix(model$encoder$fc2$weight$to(device="cpu")$detach())
  b2 <- as.numeric(model$encoder$fc2$bias$to(device="cpu")$detach())
  W3 <- as.numeric(model$encoder$fc3$weight$to(device="cpu")$detach())
  b3 <- as.numeric(model$encoder$fc3$bias$to(device="cpu")$detach())

  # W1: shape (16, n_bands)
  colnames(W1) <- BAND_NAMES
  rownames(W1) <- paste0("h1[", seq_len(nrow(W1)), "]")

  cat("**Layer 1 weights** W1 (16 × n_bands) and bias b1:\n\n")
  print(kable(round(cbind(W1, bias = b1), 4), format = "html") %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE, font_size = 10))

  # W2: shape (8, 16)
  rownames(W2) <- paste0("h2[", seq_len(nrow(W2)), "]")
  colnames(W2) <- paste0("h1[", seq_len(ncol(W2)), "]")

  cat("\n\n**Layer 2 weights** W2 (8 × 16) and bias b2:\n\n")
  print(kable(round(cbind(W2, bias = b2), 4), format = "html") %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE, font_size = 10))

  # W3: shape (1, 8) → output weights
  w3_df <- tibble(node = paste0("h2[", seq_along(W3), "]"),
                  weight = round(W3, 4))
  cat("\n\n**Layer 3 weights** w3 (1 × 8) and bias b3 =", round(b3, 4), ":\n\n")
  print(kable(w3_df, format = "html",
              col.names = c("Hidden node h2","Output weight")) %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE, font_size = 10))

  # ── Effective linear sensitivity (first-order Taylor approx) ───────────
  # Compute mean input, propagate through network, get gradient
  df_trait  <- model_data %>% filter(!is.na(.data[[trait]]))
  mean_bands <- colMeans(df_trait[, BAND_NAMES])
  bands_t    <- torch_tensor(matrix(mean_bands, nrow = 1),
                             dtype = torch_float(), requires_grad = TRUE)
  vi_val <- model$encoder(bands_t)
  vi_val$backward()
  sens <- as.numeric(bands_t$grad$to(device="cpu"))
  sens_df <- tibble(Band = BAND_NAMES,
                    `dVI/dBand (at mean)` = round(sens, 4),
                    `|Sensitivity|`       = round(abs(sens), 4)) %>%
    arrange(desc(`|Sensitivity|`))

  cat("\n\n**Linearised sensitivity** (gradient at mean band values):\n\n")
  print(kable(sens_df, format = "html") %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE))

  # ── Approximate explicit formula (top 3 bands, linear approx) ──────────
  top3 <- sens_df$Band[1:min(3, nrow(sens_df))]
  coef_str <- paste(
    sprintf("%+.4f × %s", sens_df$`dVI/dBand (at mean)`[match(top3, sens_df$Band)], top3),
    collapse = " "
  )
  cat(sprintf(
    "\n\n**Approximate linear formula** (linearised around band means):\n\n$$\\text{SynVI}_{%s} \\approx %.4f %s + \\text{const}$$\n\n",
    trait, 0, coef_str
  ))
}
```

# Comparison: Synthetic VI vs. Standard VIs

For each standard VI and trait the same LOGO scheme is used (train on 4 groups,
test on 5th) with a simple linear regression baseline.

```{r standard-vi-cv}
# LOGO linear regression benchmark for each standard VI × trait
linreg_logo_cv <- function(data, vi, trait) {
  map_dfr(group_levels_all, function(held_out) {
    tr <- data %>% filter(group != held_out,
                          !is.na(.data[[vi]]), !is.na(.data[[trait]]))
    te <- data %>% filter(group == held_out,
                          !is.na(.data[[vi]]), !is.na(.data[[trait]]))
    if (nrow(tr) < 10 || nrow(te) < 2) return(NULL)
    fit  <- lm(as.formula(paste(trait, "~", vi)), data = tr)
    pred <- predict(fit, newdata = te)
    tibble(held_out = held_out,
           r = cor(pred, te[[trait]], use = "complete.obs"))
  }) %>%
  summarise(mean_r = mean(r, na.rm = TRUE)) %>%
  pull(mean_r)
}

if (length(vi_cols) > 0) {
  cat("Computing standard VI baselines for", length(vi_cols), "VIs ×",
      length(usable_traits), "traits ...\n")
  std_vi_results <- expand_grid(vi = vi_cols, trait = usable_traits) %>%
    mutate(mean_r = map2_dbl(vi, trait,
                             ~linreg_logo_cv(model_data, .x, .y)))

  best_std_vi <- std_vi_results %>%
    group_by(trait) %>%
    slice_max(mean_r, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    rename(best_vi = vi, std_r = mean_r)

  comparison <- loeo_summary %>%
    select(trait, syn_r = mean_r) %>%
    left_join(best_std_vi, by = "trait") %>%
    mutate(improvement = syn_r - std_r)

  comparison %>%
    arrange(desc(improvement)) %>%
    kable(digits = 3,
          col.names = c("Trait","DNN Synthetic VI r","Best std VI","Std VI r","Δr"),
          caption = "DNN Synthetic VI vs. best standard VI (LOEO mean Pearson r)") %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE) %>%
    column_spec(5,
                bold  = TRUE,
                color = ifelse(comparison$improvement > 0, "green", "red"))
}
```

```{r comparison-plot, fig.cap="Prediction accuracy: Synthetic VI vs. best standard VI"}
if (length(vi_cols) > 0 && exists("comparison")) {
  comparison %>%
    pivot_longer(c(syn_r, std_r),
                 names_to = "type", values_to = "r") %>%
    mutate(type = recode(type, syn_r = "Synthetic VI (DNN)",
                               std_r = "Best standard VI (linear)"),
           trait = reorder(trait, ifelse(type == "Synthetic VI (DNN)", r, NA),
                           mean, na.rm = TRUE)) %>%
    ggplot(aes(x = r, y = trait, colour = type, shape = type)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(size = 4, alpha = 0.9) +
    geom_line(aes(group = trait), colour = "grey60", linewidth = 0.5) +
    scale_colour_manual(values = c("Synthetic VI (DNN)"    = "#e34a33",
                                   "Best standard VI (linear)" = "#3182bd")) +
    labs(x = "Mean LOEO r", y = NULL,
         title  = "Prediction accuracy: DNN Synthetic VI vs. Standard VIs (LOEO)",
         colour = NULL, shape = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

# Synthetic VI Correlations with Standard VIs

```{r synvi-stdvi-corr, fig.cap="Correlation between Synthetic VIs (per-trait) and standard VIs"}
if (length(vi_cols) > 0 && nrow(synvi_rows) > 0) {
  # Join synthetic VI values back to model_data for standard VI columns
  synvi_wide <- synvi_rows %>%
    select(Env, Timepoint, group, trait, SynVI) %>%
    pivot_wider(names_from = trait, values_from = SynVI,
                names_prefix = "SynVI_",
                values_fn = mean)   # average if >1 row per cell

  stdvi_sub <- model_data %>%
    select(Env, Timepoint, group, all_of(vi_cols)) %>%
    group_by(Env, Timepoint, group) %>%
    summarise(across(all_of(vi_cols), mean, na.rm = TRUE), .groups = "drop")

  joined <- left_join(synvi_wide, stdvi_sub, by = c("Env","Timepoint","group"))

  synvi_trait_cols <- names(joined)[startsWith(names(joined), "SynVI_")]
  if (length(synvi_trait_cols) > 0 && length(vi_cols) > 0) {
    cor_mat  <- cor(joined[, c(vi_cols, synvi_trait_cols)],
                    use = "pairwise.complete.obs")
    sub_cor  <- cor_mat[vi_cols, synvi_trait_cols, drop = FALSE]
    colnames(sub_cor) <- str_remove(colnames(sub_cor), "SynVI_")
    corrplot(sub_cor, method = "color",
             col  = colorRampPalette(c("#2166AC","white","#B2182B"))(200),
             tl.cex = 0.8, cl.cex = 0.7,
             addCoef.col = "black", number.cex = 0.65,
             title = "r(Standard VI, Synthetic VI)", mar = c(0,0,2,0))
  }
}
```

# Summary

```{r summary-table}
final_summary <- loeo_summary %>%
  left_join(
    wenv_summary %>% select(trait, wenv_r = mean_r),
    by = "trait"
  ) %>%
  left_join(
    sensitivity_list %>%
      group_by(trait) %>%
      slice_max(importance, n = 1, with_ties = FALSE) %>%
      select(trait, top_band = band),
    by = "trait"
  ) %>%
  arrange(desc(mean_r))

final_summary %>%
  kable(digits = 3,
        col.names = c("Trait","LOEO r","SD r","LOEO RMSE","N folds",
                      "Within-env r","Top band"),
        caption = "Summary: DNN Synthetic VI – LOEO and within-environment CV accuracy with dominant spectral band") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r session-info}
sessionInfo()
```

